Concurrent Programming in Harmony

### 2021-07-30 book/paper.tex
Latest

Modern databases support multiple clients concurrently accessing the data
stored in them.  Making matters more complicated, they store data on disk, but
we will ignore this in this book.
The complication we address here is that databases may be \emph{sharded},
where different parts of the data are stored on different servers with
different disks.  The different servers may even be under different
authoritive domains, such as multiple banks maintaining the accounts of their
clients.

In database terminology, a \emph{transaction} is an operation on the database.
The operation can be quite complex.  A transaction should be executed with at
least the following two properties:
\begin{itemize}
\item \emph{all-or-nothing}: a transaction should either complete, or it should
be a no-op.  It should never partially execute and then give up because of
some kind of error or something.  Database
people call this \emph{atomicity}, but it is not the same kind of atomicity that
we have been discussing in this book.
\item \emph{serialized}: any two concurrent transactions should appear to
execute in some order.  Database people call this \emph{isolation}: one transaction
should not be able to witness the intermediate state of another transaction
in execution.
\end{itemize}

We will use as an example a distributed database that maintains accounts.
For simplicity we will assume a collection of banks, each maintaining a
single account.  There are two kinds of transactions: \emph{transfer}
(similar to \autoref{ex:bank}) and \emph{inquiry}.  In this example, a transfer
is a transaction that moves some balance between two accounts.
An inquiry is a transaction over all accounts and checks that the sum of the
balances across the accounts remains the same.

@@

The Two-Phase Commit protocol~\cite{Gray78} is a protocol that can be
used to implement transactions across multiple database servers---banks
in this case.
Each transaction has a \emph{coordinator} that sends a \texttt{PREPARE}
message to each of the servers involved in the transaction, asking them
to prepare to commit to a transaction.  A server can either respond with
\texttt{YES} if it is ready to commit and will avoid doing anything
that might jeopardize this (like committing a conflicting transaction),
or \texttt{NO} if it does not want to participate in the transaction.
If all servers respond with \texttt{YES}, then the coordinator can
decide to \emph{commit} the transaction.  Otherwise the coordinator
must decide to \emph{abort} the transaction.  In the second phase, the
servers that responded with \texttt{YES} (if any) must be notified in
either case to inform them of the decision.

>>>>

Modern databases support multiple clients concurrently accessing the data
stored in them.  Making matters more complicated, they store data on disk, but
we will ignore that in this book.
The complication we address here is that databases may be \emph{sharded},
where different parts of the data are stored on different servers.
The different servers may even be under different
authoritive domains, such as multiple banks maintaining the accounts of their
clients.

In database terminology, a \emph{transaction} is an operation on a database.
The operation can be quite complex, and the execution of a transaction
should have at least the following two properties:
\begin{itemize}
\item \emph{all-or-nothing}: a transaction should either complete, or it should
be a no-op.  It should never partially execute and then give up because of
some kind of error or something.  Database
people call this \emph{atomicity}, but it is not the same kind of atomicity that
we have been discussing in this book.
\item \emph{serialized}: any two concurrent transactions should appear to
execute in some order.  Database people call this \emph{isolation}: one transaction
should not be able to witness the intermediate state of another transaction
in execution.
\end{itemize}

We will use as an example a distributed database that maintains accounts.
For simplicity we will model this as a collection of banks, each maintaining a
single account.  There are two kinds of transactions: \emph{transfer}
(similar to \autoref{ex:bank}) and \emph{inquiry}.  In this example, a transfer
is a transaction that moves some funds between two accounts.
An inquiry is a transaction over all accounts and checks that the sum of the
balances across the accounts remains the same.

@@

The Two-Phase Commit protocol~\cite{Gray78} is a protocol that can be
used to implement transactions across multiple database servers---banks
in this case.
Each transaction has a \emph{coordinator} that sends a \texttt{PREPARE}
message to each of the servers involved in the transaction, asking them
to prepare to commit to their part in a particular transaction.
A server can either respond with
\texttt{YES} if it is ready to commit and will avoid doing anything
that might jeopardize this (like committing a conflicting transaction),
or \texttt{NO} if it does not want to participate in the transaction.
If all servers respond with \texttt{YES}, then the coordinator can
decide to \emph{commit} the transaction.  Otherwise the coordinator
must decide to \emph{abort} the transaction.  In the second phase, the
servers that responded with \texttt{YES} (if any) must be notified
to inform them of the coordinator's decision.

Different transactions can have different coordinators.  In our
implementation, each bank and each coordinator is a thread.
\autoref{fig:2pc1} shows the code for a bank.
The state of a bank consists of the following local variables:
\begin{itemize}
\item \textit{self}: the bank's identifier;
\item \textit{balance}: the current balance in the account;
\item \textit{status}: either \texttt{.idle} or
$(\mathtt{.busy}, \mathit{coordinator}, \mathit{oldbalance})$,
where \textit{coordinator} is the identifier of the coordinator
of a transaction and \textit{oldbalance} is the balance
just before the transaction started;
\item \textit{received}: the set of messages received and handled so far.
\end{itemize}

\noindent
Messages sent to a bank have the following dictionary format:
$\{ \mathtt{.dst}, \mathtt{.src}, \mathtt{.request}, \mathtt{.amount} \}$:
\begin{itemize}
\item \texttt{.dst}: identifier of the bank;
\item \texttt{.src}: identifier of the coordinator that sent the message;
\item \texttt{.request}: request type, which is either
\texttt{.withdraw}, \texttt{.deposit}, \texttt{.commit}, or \texttt{.abort};
\item \texttt{.amount}: amount to withdraw or deposit.
\end{itemize}

\noindent
A bank waits for a message destined to itself that it has not yet received.
In case of a withdrawal when the bank is idle and there are sufficient
funds, the bank changes its status to \texttt{.busy} and keeps track of
who to coordinator is
